---
title: "MGPs for univariate spatial gridded data"
author: "M Peruzzi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MGPs for univariate spatial gridded data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = '40%', fig.align="center",
  message=FALSE
)
```

This is the simplest case for Meshed GPs. We start by generating some data

```{r}
library(magrittr)
library(dplyr)
library(ggplot2)
library(meshed)
set.seed(2021)

SS <- 50 # coord values for jth dimension 
dd <- 2 # spatial dimension
n <- SS^2 # number of locations
p <- 3 # number of covariates

xlocs <- seq(0, 1, length.out=SS)
coords <- expand.grid(list(xlocs, xlocs))

sigmasq <- 1.5
nu <- 0.5
phi <- 10
tausq <- .1

# covariance at coordinates
CC <- sigmasq * exp(-phi * as.matrix(dist(coords)))
# cholesky of covariance matrix
LC <- t(chol(CC))
# spatial latent effects are a GP
w <- LC %*% rnorm(n)
# measurement errors
eps <- tausq^.5 * rnorm(n)

# covariates and coefficients 
X <- matrix(rnorm(n*p), ncol=p)
Beta <- matrix(rnorm(p), ncol=1)

# univariate outcome, fully observed
y_full <- X %*% Beta + w + eps

# now introduce some NA values in the outcomes
y <- y_full %>% matrix(ncol=1)
y[sample(1:n, n/5, replace=FALSE), ] <- NA

simdata <- coords %>%
  cbind(data.frame(Outcome_full= y_full, 
                   Outcome_obs = y,
                   w = w)) 

simdata %>% ggplot(aes(Var1, Var2, fill=w)) +
    geom_raster() + 
    scale_fill_viridis_c() +
  theme_minimal() + ggtitle("w: Latent GP")
simdata %>% ggplot(aes(Var1, Var2, fill=y)) +
    geom_raster() + 
    scale_fill_viridis_c() + 
  theme_minimal() + ggtitle("y: Observed outcomes")
```
We now fit the following spatial regression model 
$$ y = X \beta + w + \epsilon, $$
where $w \sim MGP$ are spatial random effects, and $\epsilon \sim N(0, \tau^2)$. For spatial data, an exponential covariance function is used by default: $C(h) = \sigma^2 \exp( -\phi h )$ where $h$ is the spatial distance.

The prior for the spatial decay $\phi$ is $U[l,u]$ and the values of $l$ and $u$ must be specified. We choose $l=1$, $u=30$ for this dataset.^[`spmeshed` implements a model which can be interpreted as assigning $\sigma^2$ a folded-t via parameter expansion if `settings$ps=TRUE`, or an inverse Gamma with parameters $a=2$ and $b=1$ if `settings$ps=FALSE`, which cannot be changed at this time. $\tau^2$ is assigned an exponential prior.] 

Setting `verbose` tells `spmeshed` how many intermediate messages to output while running MCMC. We set up MCMC to run for 3000 iterations, of which we discard the first third. We use the argument `block_size=25` to specify the coarseness of domain partitioning. In this case, the same result can be achieved by setting `axis_partition=c(10, 10)`. 

Finally, we note that `NA` values for the outcome are OK since the full dataset is on a grid. `spmeshed` will figure this out and use settings optimized for partly observed lattices, and automatically predict the outcomes at missing locations. On the other hand, `X` values are assumed to not be missing. 

```{r}
mcmc_keep <- 200 # too small! this is just a vignette.
mcmc_burn <- 400
mcmc_thin <- 2

mesh_total_time <- system.time({
  meshout <- spmeshed(y, X, coords,
                      #axis_partition=c(10,10), #same as block_size=25
                      block_size = 25,
                      n_samples = mcmc_keep, n_burn = mcmc_burn, n_thin = mcmc_thin, 
                      n_threads = 2,
                      verbose = 0,
                      prior=list(phi=c(1,30))
  )})
```
We can now do some postprocessing of the results. We extract posterior marginal summaries for $\sigma^2$, $\phi$, $\tau^2$, and $\beta_2$. The model that `spmeshed` targets is a slight reparametrization of the above:^[At its core, `spmeshed` implements the spatial factor model $Y(s) = X(s)\beta + \Lambda v(s) + \epsilon(s)$ where $w(s) = \Lambda v(s)$ is modeled via linear coregionalization.]
$$ y = X \beta + \lambda w + \epsilon, $$
where $w\sim MGP$ has unitary variance. This model is equivalent to the previous one and in fact we find $\sigma^2=\lambda^2$. 

```{r}
summary(meshout$lambda_mcmc[1,1,]^2)
summary(meshout$theta_mcmc[1,1,])
summary(meshout$tausq_mcmc[1,])
summary(meshout$beta_mcmc[1,1,])
```
We proceed to plot predictions across the domain along with the recovered latent effects.
```{r}
# process means
wmesh <- data.frame(w_mgp = meshout$w_mcmc %>% summary_list_mean())
# predictions
ymesh <- data.frame(y_mgp = meshout$yhat_mcmc %>% summary_list_mean())

outdf <- meshout$coordsdata %>% 
  cbind(wmesh, ymesh) %>%
  left_join(simdata)

outdf %>% 
    ggplot(aes(Var1, Var2, fill=w_mgp)) +
    geom_raster() +
    scale_fill_viridis_c() +
    theme_minimal() + ggtitle("w: recovered")

outdf %>% 
    ggplot(aes(Var1, Var2, fill=y_mgp)) +
    geom_raster() +
    scale_fill_viridis_c() +
    theme_minimal() + ggtitle("y: predictions")
```